{\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of finite-dimensional and affine Lie algebras

### 2011-05-11 doc/paper.tex
More work on text

% \biboptions{}
\newcommand{\go}{\stackrel{\circ }{\mathfrak{g}}}
\newcommand{\ao}{\stackrel{\circ }{\mathfrak{a}}}
\newcommand{\co}[1]{\stackrel{\circ }{#1}}
\newcommand{\pia}{\pi_{\mathfrak{a}}}
\newcommand{\piab}{\pi_{\mathfrak{a}_{\bot}}}
\newcommand{\gf}{\mathfrak{g}}
\newcommand{\af}{\mathfrak{a}}
\newcommand{\afb}{\mathfrak{a}_{\bot}}
\newcommand{\hf}{\mathfrak{h}}
\newcommand{\hfg}{\hf_{\gf}}
\newcommand{\hfb}{\mathfrak{h}_{\bot}}

@@

For an arbitrary representation it is possible to diagonalize Cartan generators $H^{i}$ simultaneously by special choice of basis $\left\{\left|j\right>\right\}$ in $V$:
\begin{equation}
  \label{eq:3}
  H^{i}\left|j\right>=\nu_{j}^{i}\left|j\right>
\end{equation}
Eigenvalues of Cartan generators on element of basis $\left|j\right>$ are given by weight $\nu_{j}\in \hfg^{*}: \nu_{j}(H^{i})=\nu_{j}^{i}$, correspondingly vector $\left|j\right>$ is called weight vector of $\nu_{j}$.

@@

The structure of module is determined by the set of weights since the action of generators $E^{\alpha}$ on weight vectors is
\begin{equation}
  \label{eq:5}
  E^{\alpha}\left|\nu\right> \propto \left|\nu+\alpha\right>
\end{equation}

In the root system $\Delta$ exist simple roots $\left\{\alpha_{i}\right\}$, such that $\alpha=\sum_{i}n_{i}\alpha_{i},\; \forall\alpha\in \Delta$. In case of finite-dimensional Lie algebra $\gf$ $i=1,\dots,r,\quad r=\mathrm{rank}(\gf)$. Although the full set of roots $\Delta$ is infinite for affine Lie algebra $\hat\gf$ the number of simple roots is finite and are denoted by $\alpha_{0},\dots \alpha_{r}$ where $r=\mathrm{rank}(\gf)$. Roots system can be reconstructed from the set of simple roots with the use of Weyl group $W_{\gf}$ which is generated by reflections $\{s_{i}:\hfg^{*}\to\hfg^{*}\}$ corresponding to simple roots $\{\alpha_{i}\}$:
\begin{equation}
  \label{eq:8}
  s_{i}\cdot\lambda=\lambda-\frac{2(\alpha_{i},\lambda)}{(\alpha_{i},\alpha_{i})}\alpha_{i}
\end{equation}
Weyl groups are finite for finite-dimensional Lie algebras and finitely-generated for affine Lie algebras. In the latter case of $\hat\gf$ Weyl group can be presented as semidirect product of Weyl group $W_{\gf}$ of $\gf$ and translations by the roots of $\hat \gf$.

@@

>>>>

% \biboptions{}
\newcommand{\go}{\stackrel{\circ }{\mathfrak{g}}}
\newcommand{\ao}{\stackrel{\circ }{\mathfrak{a}}}
\newcommand{\co}[1]{\stackrel{\circ }{#1}}
\newcommand{\pia}{\pi_{\mathfrak{a}}}
\newcommand{\piab}{\pi_{\mathfrak{a}_{\bot}}}
\newcommand{\gf}{\mathfrak{g}}
\newcommand{\af}{\mathfrak{a}}
\newcommand{\bff}{\mathfrak{b}}
\newcommand{\afb}{\mathfrak{a}_{\bot}}
\newcommand{\hf}{\mathfrak{h}}
\newcommand{\hfg}{\hf_{\gf}}
\newcommand{\hfb}{\mathfrak{h}_{\bot}}

@@

For an arbitrary representation it is possible to diagonalize Cartan generators $H^{i}$ simultaneously by special choice of basis $\{v_{j}\}$ in $V$:
\begin{equation}
  \label{eq:3}
  H^{i}\cdot v_{j}=\nu_{j}^{i}v_{j}
\end{equation}
Eigenvalues of Cartan generators on element of basis $v_{j}$ are given by weight $\nu_{j}\in \hfg^{*}: \nu_{j}(H^{i})=\nu_{j}^{i}$. Vector $v\in V$ is called weight vector of weight $\lambda$ if $H v=\lambda_{j}(H)v,\; \forall H\in \hf$ , weight subspace consists of all weight vectors $V_{\lambda}=\{v\in V: H v=\lambda_{j}(H)v,\; \forall H\in \hf\}$. Weight multiplicity $m_{\lambda}=\mathrm{mult}(\lambda)=\mathrm{dim} V_{\lambda}$ is the dimension of weight subspace.

The structure of module is determined by the set of weights since the action of generators $E^{\alpha}$ on weight vectors is
\begin{equation}
  \label{eq:5}
  E^{\alpha}\cdot v_{\lambda} \propto v_{\lambda+\alpha}
\end{equation}
It can be encoded by the formal character of module
\begin{equation}
  \label{eq:10}
  \mathrm{ch}V=\sum_{\lambda}m_{\lambda} e^{\lambda}
\end{equation}
Character  $\mathrm{ch}V\in \mathcal{E}$. is the element of algebra $\mathcal{E}$ generated by formal exponents of weights.

@@

In the root system $\Delta$ exist simple roots $\left\{\alpha_{i}\right\}$, such that $\alpha=\sum_{i}n_{i}\alpha_{i},\; \forall\alpha\in \Delta$. In case of finite-dimensional Lie algebra $\gf$ $i=1,\dots,r,\quad r=\mathrm{rank}(\gf)$. Although the full set of roots $\Delta$ is infinite for affine Lie algebra $\hat\gf$ the number of simple roots is finite and are denoted by $\alpha_{0},\dots \alpha_{r}$ where $r=\mathrm{rank}(\gf)$. Roots system can be reconstructed from the set of simple roots with the use of Weyl group $W_{\gf}$ which is generated by reflections $\{s_{i}:\hfg^{*}\to\hfg^{*}\}$ corresponding to simple roots $\{\alpha_{i}\}$:
\begin{equation}
  \label{eq:8}
  s_{i}\cdot\lambda=\lambda-\frac{2(\alpha_{i},\lambda)}{(\alpha_{i},\alpha_{i})}\alpha_{i}
\end{equation}
Note that root multiplicity $\mathrm{mult}(\alpha)$ for affine Lie algebra can be greater than one. Weyl vector is denoted by $\rho=\frac{1}{2}\sum_{\alpha\in \Delta^{+}}\alpha$.

Weyl groups are finite for finite-dimensional Lie algebras and finitely-generated for affine Lie algebras. In the latter case of $\hat\gf$ Weyl group can be presented as semidirect product of Weyl group $W_{\gf}$ of $\gf$ and translations by the roots of $\hat \gf$.

@@

We consider $\gf$-modules of the category $\mathcal{O}$, i.e. such $V$ that $V=\bigoplus_{\lambda\in \hfg^{*}} V_{\lambda}$, where each $V_{\lambda}$ is finite-dimensional and there exists finite set $\lambda_{1},\dots \lambda_{s}$ such that $V_{\lambda}\neq 0 \Rightarrow \lambda=\lambda_{i}-\sum_{k=1,\dots, r} n_{k}\alpha_{k}$, where $n_{k}\in \mathbb{Z}_{+}$ (See \cite{humphreys2008representations}, \cite{carter2005lie}).

Highest weight module $V^{\mu}$ contains one highest weight $\mu$, all other weights are obtained by the subtraction of linear combination of simple roots $\lambda=\mu-n_{1}\alpha_{1}-\dots-n_{r}\alpha_{r},\; n_{k}\in \mathbb{Z}_{+}$.

The simplest type of highest weight modules is Verma module $M^{\mu}$ which can be defined as the space $U(\gf)\underset{U(\bff_{+})}{\otimes} D_{\lambda}(\bff_{+})$, where $\underset{U(\bff_{+})}{\otimes}$ means that the action of elements of $U(\bff_{+})$ ``falls through'' left part of tensor product onto the right part. Here $\bff_{+}$ is Borel sub-algebra, $D_{\lambda}(\bff_{+})$ is a special representation of $\bff_{+}$, such that $D(E^{\alpha})=0,\; D(h)=\lambda(h)$ for any positive root $\alpha$.
Elements of $\gf$ act from the left, so we should expand element of $\gf$ in some basis, and then commute all the elements of $\bff_{+}$ to the right, so that they can act on space $D_{\lambda}(\bff_{+})$.

Weight multiplicities of Verma module can be found from Weyl character formula
\begin{equation}
  \label{eq:11}
  \mathrm{ch} M^{\mu}=\frac{e^{\mu}}{\prod_{\alpha\in \Delta^{+}} \left( 1-e^{-\alpha}\right)^{\mathrm{mult}(\alpha)}}=\frac{e^{\mu}}{\sum_{w\in W} \epsilon(w) e^{w\rho-\rho}}
\end{equation}
Here we have used Weyl denominator identity
\begin{equation}
  \label{eq:12}
  \prod_{\alpha\in \Delta^{+}} \left( 1-e^{-\alpha}\right)^{\mathrm{mult}(\alpha)}=\sum_{w\in W} \epsilon(w) e^{w\rho-\rho},
\end{equation}
and $\epsilon \left( w\right) :=\det \left( w\right)$ is also equal to the parity of number of Weyl reflections generating $w$.

Verma module $M^{\mu}$ has unique maximal submodule and unique nontrivial simple quotient $L^{\mu}$ which is irreducible highest weight module.

Weyl character formula for irreducible highest weight modules is
\begin{equation}
  \label{eq:13}
  \mathrm{ch} L^{\mu}=\frac{\sum_{w\in W} \epsilon(w) e^{w(\mu+\rho)-\rho}}{\sum_{w\in W}\epsilon(w) e^{w\rho-\rho}}
\end{equation}
So character of irreducible highest weight module can be seen as the combination of characters of Verma modules, this is connected with Bernstein-Gelfand-Gelfand resolution (\cite{bernstein1976category,bernstein1971structure}, see also \cite{humphreys2008representations}).

We can use Weyl character formula to obtain recurrent relation on weight multiplicities which can be used for calculations \cite{il2010folded,kulish4sfa}:

This formula is a particular case of more general formula related to branching which will discuss soon.

Irreducible highest weight modules, BGG resolution, anomalous element.
